---
title: "Global CO₂ Emissions: Does GDP Predict Emissions and Does Carbon Tax Help?"
author: "Isuri"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    code_folding: show
    df_print: paged
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6,
  fig.align  = "center"
)
```
```{r libraries}
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(broom)
library(knitr)
library(kableExtra)
```

---

# 1. Introduction

This case study investigates two related research questions about global CO₂ emissions:

> **Q1:** Does GDP per capita predict CO₂ emissions per capita across countries?

> **Q2:** Does having a carbon tax affect CO₂ emissions in top emitting countries?

Understanding the relationship between economic development, carbon pricing policy,
and emissions is critical for designing effective climate policy. A naive analysis
might suggest carbon tax increases emissions: this study demonstrates why
controlling for confounding variables matters.

## 1.1 Data Sources

| Dataset | Source | Coverage |
|---|---|---|
| CO₂ & Greenhouse Gas Emissions | Our World in Data | 164 countries, 2000–2024 |
| Carbon Pricing Instrument Coverage | World Bank | 164 countries, 1989–2024 |

---

# 2. Data Loading & Cleaning

## 2.1 Load raw data
```{r load-data}
co2EmissionData <- read_csv("data/raw/co2-emission.csv",
                             show_col_types = FALSE)
carbonTaxData   <- read_csv("data/raw/carbon-tax-data.csv",
                             show_col_types = FALSE)

cat("CO2 dataset:", nrow(co2EmissionData), "rows x",
    ncol(co2EmissionData), "columns\n")
cat("Carbon tax dataset:", nrow(carbonTaxData), "rows x",
    ncol(carbonTaxData), "columns\n")
```

## 2.2 Clean carbon tax data
```{r clean-tax}
carbonTaxData <- carbonTaxData %>%
  rename(
    country        = Entity,
    iso_code       = Code,
    year           = Year,
    has_carbon_tax = `Covered by tax instrument in at least one sector`
  ) %>%
  mutate(
    carbon_tax = ifelse(has_carbon_tax == "No carbon tax",
                        FALSE, TRUE)
  ) %>%
  select(country, iso_code, year, carbon_tax) %>%
  filter(year >= 2000 & year <= 2024)

cat("Unique carbon_tax values:",
    paste(unique(carbonTaxData$carbon_tax), collapse = ", "), "\n")
```

## 2.3 Clean CO₂ data
```{r clean-co2}
co2EmissionData <- co2EmissionData %>%
  filter(!is.na(iso_code),
         year >= 2000 & year <= 2024)

co2EmissionDataCleaned <- co2EmissionData %>%
  select(country, iso_code, year,
         gdp, population, co2,
         co2_per_capita, share_global_co2,
         primary_energy_consumption)

co2EmissionDataCleaned <- co2EmissionDataCleaned %>%
  arrange(country, year) %>%
  group_by(country) %>%
  mutate(
    gdp = ifelse(is.na(gdp),
                 mean(gdp, na.rm = TRUE), gdp),
    primary_energy_consumption = ifelse(
      is.na(primary_energy_consumption),
      mean(primary_energy_consumption, na.rm = TRUE),
      primary_energy_consumption)
  ) %>%
  ungroup()

no_gdp_countries <- co2EmissionDataCleaned %>%
  group_by(country) %>%
  summarise(all_na = all(is.na(gdp)), .groups = "drop") %>%
  filter(all_na) %>%
  pull(country)

cat("Countries removed (no GDP data):",
    length(no_gdp_countries), "\n")

co2EmissionDataCleaned <- co2EmissionDataCleaned %>%
  filter(!country %in% no_gdp_countries) %>%
  mutate(gdp_per_capita = gdp / population)

cat("\nMissing values after cleaning:\n")
print(colSums(is.na(co2EmissionDataCleaned)))
```

## 2.4 Join datasets
```{r join}
co2_tax_gdp <- co2EmissionDataCleaned %>%
  inner_join(
    carbonTaxData %>% select(iso_code, year, carbon_tax),
    by = c("iso_code", "year")
  )

cat("Final dataset:", nrow(co2_tax_gdp), "rows\n")
cat("Countries:", n_distinct(co2_tax_gdp$country), "\n")
cat("Year range:",
    paste(range(co2_tax_gdp$year), collapse = " to "), "\n")
```

---

# 3. Data Aggregation
```{r aggregate}
country_avg <- co2_tax_gdp %>%
  group_by(country) %>%
  summarise(
    avg_co2_per_capita = mean(co2_per_capita,  na.rm = TRUE),
    avg_gdp_per_capita = mean(gdp_per_capita,  na.rm = TRUE),
    avg_share_global   = mean(share_global_co2, na.rm = TRUE),
    avg_energy         = mean(primary_energy_consumption,
                              na.rm = TRUE),
    has_carbon_tax     = as.logical(max(carbon_tax,
                                        na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  mutate(
    gdp_level = case_when(
      avg_gdp_per_capita > 12000 ~ "High income",
      avg_gdp_per_capita > 3000  ~ "Middle income",
      TRUE                       ~ "Low income"
    ),
    gdp_level = factor(gdp_level,
                       levels = c("Low income",
                                  "Middle income",
                                  "High income"))
  )

top_countries <- co2_tax_gdp %>%
  filter(year >= 2018 & year <= 2024) %>%
  group_by(country) %>%
  summarise(avg_share = mean(share_global_co2, na.rm = TRUE),
            .groups = "drop") %>%
  arrange(desc(avg_share)) %>%
  slice_head(n = 20) %>%
  pull(country)

cat("Dataset dimensions:", nrow(country_avg),
    "countries\n")
cat("Top 20 emitters identified\n")
```

---

# 4. Exploratory Data Analysis

## 4.1 Distribution of CO₂ per capita
```{r eda-hist}
ggplot(country_avg, aes(x = avg_co2_per_capita)) +
  geom_histogram(binwidth = 1, fill = "#47B9C6",
                 color = "white") +
  labs(
    title    = "Distribution of Average CO2 Emissions per Capita",
    subtitle = "Right-skewed: most countries emit under 5 tonnes",
    x = "Average CO2 per Capita (tonnes)",
    y = "Number of Countries"
  ) +
  theme_minimal(base_size = 12)
```

The distribution is strongly right-skewed. Most countries emit fewer than
5 tonnes per capita, while a small number of high-income and fossil-fuel-dependent
nations are significant outliers above 15 tonnes.

## 4.2 Top 20 countries by global CO₂ share
```{r eda-top20}
country_avg %>%
  filter(country %in% top_countries) %>%
  arrange(desc(avg_share_global)) %>%
  mutate(top5 = row_number() <= 5) %>%
  ggplot(aes(x = reorder(country, avg_share_global),
             y = avg_share_global, fill = top5)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE" = "#21497E", "FALSE" = "#ACE9DF"),
    guide  = "none"
  ) +
  labs(
    title    = "Top 20 Countries by Average Share of Global CO2",
    subtitle = "China and USA account for ~42% of global emissions",
    x = "Country",
    y = "Average Share of Global CO2 (%)"
  ) +
  theme_minimal(base_size = 12)
```

China (25%) and the United States (17%) dominate global emissions.
The top 5 emitters together account for over 55% of global CO₂.

## 4.3 Carbon tax adoption
```{r eda-donut}
carbon_tax_dist <- country_avg %>%
  group_by(has_carbon_tax) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(
    pct   = round(count / sum(count) * 100, 1),
    label = ifelse(has_carbon_tax,
                   "Has Carbon Tax", "No Carbon Tax")
  )

ggplot(carbon_tax_dist,
       aes(x = 2, y = pct, fill = label)) +
  geom_bar(stat = "identity", width = 1,
           color = "white", linewidth = 0.5) +
  coord_polar(theta = "y") +
  xlim(0.5, 2.5) +
  geom_text(aes(label = paste0(label, "\n", pct, "%")),
            position = position_stack(vjust = 0.5),
            fontface = "bold", size = 4) +
  scale_fill_manual(values = c(
    "Has Carbon Tax" = "#47B9C6",
    "No Carbon Tax"  = "#ACE9DF"
  )) +
  labs(title = "Carbon Tax Adoption Across Countries") +
  theme_void() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5,
                                  face = "bold",
                                  size = 14))
```

Only **17.1%** of countries have adopted a carbon tax. This limited adoption
makes cross-country comparison feasible but means the taxed group is a minority.

## 4.4 GDP per capita vs CO₂ per capita
```{r eda-scatter}
country_avg %>%
  filter(country %in% top_countries) %>%
  ggplot(aes(x = avg_gdp_per_capita,
             y = avg_co2_per_capita)) +
  geom_point(color = "#47B9C6", size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE,
              color = "#21497E", fill = "#ACE9DF",
              alpha = 0.2) +
  geom_text_repel(aes(label = country),
                  size = 3, color = "grey30") +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(
    title    = "GDP per Capita vs CO2 per Capita (Top 20 Emitters)",
    subtitle = "Strong positive linear relationship",
    x = "Average GDP per Capita (USD)",
    y = "Average CO2 per Capita (tonnes)"
  ) +
  theme_minimal(base_size = 12)
```

A clear positive relationship is visible. Notable exceptions include
Saudi Arabia (high emissions relative to GDP) and Germany/UK (lower
emissions than GDP would predict, possibly due to carbon pricing and
energy efficiency policies).

## 4.5 CO₂ per capita by country coloured by carbon tax
```{r eda-bar-tax}
country_avg %>%
  filter(country %in% top_countries) %>%
  arrange(avg_co2_per_capita) %>%
  ggplot(aes(x = reorder(country, avg_co2_per_capita),
             y = avg_co2_per_capita,
             fill = has_carbon_tax)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(
    values = c("TRUE"  = "#47B9C6",
               "FALSE" = "#ACE9DF"),
    labels = c("TRUE"  = "Has Carbon Tax",
               "FALSE" = "No Carbon Tax"),
    name   = "Carbon Tax"
  ) +
  labs(
    title    = "Average CO2 per Capita: Top 20 Emitters",
    subtitle = "Carbon tax countries not consistently lower emitters",
    x = "Country",
    y = "Average CO2 per Capita (tonnes)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

Countries with carbon tax are not consistently lower emitters:
Australia and Canada have high emissions despite carbon tax.
This confirms GDP confounding motivates the regression analysis below.

## 4.6 GDP vs CO₂ coloured by carbon tax
```{r eda-scatter-tax}
country_avg %>%
  filter(country %in% top_countries) %>%
  ggplot(aes(x = avg_gdp_per_capita,
             y = avg_co2_per_capita,
             color = has_carbon_tax)) +
  geom_point(size = 3, alpha = 0.85) +
  geom_text_repel(aes(label = country),
                  size = 2.8) +
  scale_color_manual(
    values = c("TRUE"  = "#47B9C6",
               "FALSE" = "gray50"),
    labels = c("TRUE"  = "Has Carbon Tax",
               "FALSE" = "No Carbon Tax"),
    name   = "Carbon Tax"
  ) +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(
    title    = "GDP vs CO2 per Capita: Coloured by Carbon Tax",
    subtitle = "High-GDP carbon tax countries cluster above the trend line",
    x = "Average GDP per Capita (USD)",
    y = "Average CO2 per Capita (tonnes)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

High-GDP countries with carbon tax (Germany, UK, Japan) sit *below*
where the trend line would place them: suggesting carbon tax may be
contributing to lower-than-expected emissions at high income levels.

---

# 5. Correlation Analysis

## 5.1 Q1 : GDP per capita vs CO₂ per capita
```{r cor-q1}
cor_q1 <- cor.test(
  country_avg$avg_gdp_per_capita,
  country_avg$avg_co2_per_capita,
  method = "pearson"
)

tidy(cor_q1) %>%
  select(estimate, statistic, p.value,
         conf.low, conf.high) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(
    caption = "Pearson Correlation: GDP vs CO2 per Capita",
    col.names = c("r", "t-statistic", "p-value",
                  "CI lower", "CI upper")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

There is a **strong positive correlation** between GDP per capita and
CO₂ per capita (r = `r round(cor_q1$estimate, 3)`, p < 0.001).
The 95% confidence interval confirms this is a reliable and
substantial relationship. Wealthier countries consistently emit
more CO₂ per person.

## 5.2 Q2 : Carbon tax vs CO₂ per capita
```{r cor-q2}
cor_q2 <- cor.test(
  as.numeric(country_avg$has_carbon_tax),
  country_avg$avg_co2_per_capita,
  method = "pearson"
)

tidy(cor_q2) %>%
  select(estimate, statistic, p.value,
         conf.low, conf.high) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(
    caption = "Pearson Correlation: Carbon Tax vs CO2 per Capita",
    col.names = c("r", "t-statistic", "p-value",
                  "CI lower", "CI upper")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

Without controlling for GDP, carbon tax shows a **weak positive**
correlation with CO₂ per capita
(r = `r round(cor_q2$estimate, 3)`,
p = `r round(cor_q2$p.value, 4)`).
This counterintuitive direction is explained by GDP confounding :
richer countries both adopt carbon tax and emit more per capita.
This motivates the controlled regression analysis below.

## 5.3 Correlation summary
```{r cor-summary}
data.frame(
  Relationship        = c("GDP per capita → CO2 per capita",
                          "Carbon tax → CO2 per capita"),
  r                   = c(round(cor_q1$estimate, 3),
                          round(cor_q2$estimate, 3)),
  p_value             = c(round(cor_q1$p.value, 6),
                          round(cor_q2$p.value, 4)),
  Interpretation      = c("Strong positive : significant",
                          "Weak positive : confounded by GDP")
) %>%
  kable(caption = "Correlation Analysis Summary") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

---

# 6. Regression Analysis

## 6.1 Model 1 : GDP alone
```{r reg-m1}
m1 <- lm(avg_co2_per_capita ~ avg_gdp_per_capita,
         data = country_avg)

tidy(m1) %>%
  mutate(across(where(is.numeric), ~round(., 5))) %>%
  kable(caption = "Model 1: GDP per Capita → CO2 per Capita") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )

glance(m1) %>%
  select(r.squared, adj.r.squared,
         sigma, p.value, nobs) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(caption = "Model 1 : Model Performance") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

GDP per capita alone explains
**`r round(glance(m1)$r.squared * 100, 1)`%**
of variation in CO₂ per capita. For every $10,000 increase in
GDP per capita, CO₂ per capita increases by approximately
**`r round(tidy(m1)$estimate[2] * 10000, 2)` tonnes**.

## 6.2 Model 2 : Carbon tax alone
```{r reg-m2}
m2 <- lm(avg_co2_per_capita ~ has_carbon_tax,
         data = country_avg)

tidy(m2) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(caption = "Model 2: Carbon Tax → CO2 per Capita") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

Without controlling for GDP, having a carbon tax is associated with
**+`r round(tidy(m2)$estimate[2], 2)` tonnes** more CO₂ per capita.
This is misleading : it reflects that wealthy countries adopt
carbon tax AND emit more due to their economic scale.

## 6.3 Model 3 : GDP + Carbon tax
```{r reg-m3}
m3 <- lm(avg_co2_per_capita ~ avg_gdp_per_capita + has_carbon_tax,
         data = country_avg)

tidy(m3) %>%
  mutate(across(where(is.numeric), ~round(., 5))) %>%
  kable(caption = "Model 3: GDP + Carbon Tax → CO2 per Capita") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )

glance(m3) %>%
  select(r.squared, adj.r.squared,
         sigma, p.value, nobs) %>%
  mutate(across(where(is.numeric), ~round(., 4))) %>%
  kable(caption = "Model 3 : Model Performance") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

After controlling for GDP, carbon tax is associated with
**`r round(tidy(m3)$estimate[3], 2)` tonnes less** CO₂ per capita
(p < 0.001). The sign completely flips from positive to negative :
confirming GDP as a confounding variable. Together GDP and carbon
tax explain
**`r round(glance(m3)$r.squared * 100, 1)`%**
of CO₂ variation.

## 6.4 Model comparison
```{r reg-compare}
data.frame(
  Model = c(
    "M1: GDP alone",
    "M2: Carbon tax alone",
    "M3: GDP + Carbon tax"
  ),
  R_squared = c(
    round(glance(m1)$r.squared, 3),
    round(glance(m2)$r.squared, 3),
    round(glance(m3)$r.squared, 3)
  ),
  Carbon_tax_coef = c(
    "N/A",
    paste0("+", round(tidy(m2)$estimate[2], 3)),
    round(tidy(m3)$estimate[3], 3)
  ),
  Interpretation = c(
    "GDP explains 67.8% alone",
    "Misleading positive : GDP confounded",
    "True effect after GDP control"
  )
) %>%
  kable(
    caption = "Model Comparison : Carbon Tax Coefficient",
    col.names = c("Model", "R²",
                  "Carbon tax coefficient",
                  "Interpretation")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  row_spec(3, bold = TRUE, background = "#ACE9DF")
```

## 6.5 Actual vs fitted plot
```{r reg-fitted}
country_avg %>%
  mutate(fitted = fitted(m3)) %>%
  filter(country %in% top_countries) %>%
  ggplot(aes(x = fitted,
             y = avg_co2_per_capita,
             color = has_carbon_tax,
             label = country)) +
  geom_point(size = 3) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed",
              color = "#21497E") +
  geom_text_repel(size = 2.8) +
  scale_color_manual(
    values = c("TRUE"  = "#47B9C6",
               "FALSE" = "gray50"),
    labels = c("TRUE"  = "Has Carbon Tax",
               "FALSE" = "No Carbon Tax"),
    name   = "Carbon Tax"
  ) +
  labs(
    title    = "Actual vs Fitted CO2 per Capita (Model 3)",
    subtitle = "Points close to dashed line = well predicted by model",
    x = "Fitted values",
    y = "Actual CO2 per Capita (tonnes)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

Points close to the dashed line are well predicted by the model.
Saudi Arabia sits well above : its emissions are higher than
the model predicts given its GDP, suggesting fossil fuel
dependence beyond what income alone explains.

---

# 7. Key Findings
```{r findings-table}
data.frame(
  Finding = c(
    "GDP strongly predicts CO2",
    "Carbon tax misleading without control",
    "Carbon tax reduces CO2 after GDP control",
    "Model explanatory power",
    "Confounding variable confirmed"
  ),
  Detail = c(
    "r = 0.824, p < 0.001 : strong positive relationship",
    "Without GDP: +3.59 tonnes (positive, misleading)",
    "With GDP control: -3.24 tonnes (p < 0.001)",
    "GDP + tax explains 70.8% of CO2 variation (R2 = 0.708)",
    "GDP is a confounder : direction flips when controlled"
  )
) %>%
  kable(caption = "Summary of Key Findings") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE
  ) %>%
  column_spec(1, bold = TRUE, color = "#21497E")
```

### Q1 Answer
GDP per capita is a **strong and statistically significant predictor**
of CO₂ emissions per capita (r = 0.824, R² = 0.678). For every
$10,000 increase in GDP per capita, a country emits approximately
3.2 additional tonnes of CO₂ per person per year.

### Q2 Answer
Carbon tax **does reduce CO₂ emissions**, but only after controlling
for GDP. Without the control, it appears to increase emissions :
a classic example of confounding. After accounting for GDP,
carbon tax is associated with 3.24 fewer tonnes of CO₂ per capita
(p < 0.001).

---

# 8. Limitations
```{r limitations}
data.frame(
  Limitation = c(
    "Small sample",
    "Binary tax variable",
    "Cross-sectional aggregation",
    "Omitted variables",
    "Causality"
  ),
  Detail = c(
    "Only 164 countries: limits statistical power",
    "Carbon tax is yes/no: does not capture tax rate or coverage",
    "Country averages mask year-to-year variation",
    "Energy mix, industrial structure, and policy stringency not included",
    "Regression shows association, not causation"
  )
) %>%
  kable(caption = "Study Limitations") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE
  )
```

---

# 9. References

- **Our World in Data** - CO₂ and Greenhouse Gas Emissions dataset.
  Hannah Ritchie, Max Roser and Pablo Rosado (2020).
  [ourworldindata.org/co2-and-greenhouse-gas-emissions](https://ourworldindata.org/co2-and-greenhouse-gas-emissions)

- **World Bank** - Carbon Pricing Dashboard.
  Carbon tax instrument coverage by country and year.
  [carbonpricingdashboard.worldbank.org](https://carbonpricingdashboard.worldbank.org)

- **R Core Team** (2024). R: A language and environment for
  statistical computing. R Foundation for Statistical Computing,
  Vienna, Austria.

- **Wickham H. et al.** (2019). Welcome to the tidyverse.
  Journal of Open Source Software, 4(43), 1686.

---
